Optimal Algorithms for Dynamic Balanced Graph Partitioning

### 2020-05-14 paper/disc20clustering.tex
some comments done

	We proceed in similar steps by joining the current pivot node to a ground set of the same origin residing in a different cluster.
	Consider the step $i$,
	when the isolated node $x_i$ is collocated with $B$.
	 We issue a request between $x_i$ and some node in $C_i$,
	 where $C_i$ is the largest ground set
	 s.t.~$I(C_i) = I(x_i), C_i \neq \{x_0,y_0\}$ \maciek{Ground sets merges instead of requests}.
%	 only if  a feasible partition exists
%	 that collocates all nodes in $\{x_i\} \cup C_i $.
	Any feasible partition replaces $x_i$ with some isolated node $x_{i+1}$,
	as the new ground set $\{x_i\} \cup C_i $  may not be ever split.
%	Otherwise,
%	no such feasible partition exists and
%	we terminate the process.
	We terminate the process once the number of remaining isolated nodes 
	is less than $\ell+3$.
	At each step $i$,
	the number of isolated nodes decrease by one or two \maciek{Is this clear why? I am not, sure, take a look and decide. Possibly we need to say about two singletons merge}.
	Therefore, once the process terminates,
	there are at least $\ell+1$ isolated nodes left.

@@

We maintain subsets of frequently communicating nodes as \emph{components}.
Initially,
each node constitutes a single-node component which we refer to as a \emph{singleton} component, 
and the contained node is an \emph{isolated} node \maciek{unclear: isolated = contained?}.
We define larger components in terms of smaller components.
Concretely,
given a (sub)sequence of requests,
two components $C_1$ and $C_2$
 \emph{merge} into one component as soon as
 for some pair of nodes $v_1 \in C_1$ and $v_2 \in C_2$,
the frequency of requests $\{v_1,v_2\}$ reaches a certain threshold through the sequence.
We keep all nodes of a~component always collocated in the same cluster,
i.e., when we  move a~node,
we move the whole component that contains it.
A~partition that maintains this invariant is a~\emph{component respecting} partition \maciek{wrong use of "this". Better to define component respecting, and only then say that ALG only uses such partitions (but it is also implied in the learning model...)}.
%Intuitively, a component preserves the history of requests between its nodes.

@@

The algorithm in this section and the algorithm for $k=3$ (cf. Section~\ref{sec:k3}) are modified versions of the algorithm DET from~\cite{repartition-disc}.
The difference is in the choice of partition after a~component merge.
In DET, the partition was arbitrary.
In the algorithm for the learning model,
we choose a~component respecting partition closest to the initial partition.
In the algorithm from Section~\ref{sec:k3}, we choose the partition closest to the current partition (a repartition of minimum cost).
A generic sketch of these two algorithms can be found in the appendix (Algorithm \ref{alg:ppl}) \maciek{We don't have the generic. Also in the next paragraph you refer to the code}.

\noindent
\textbf{Perfect Partition Learner algorithm.}
\maciek{There should be a short description of the algorithm: what steps it takes, how it treats the input. Instead, we have PPL's properties here, and the algorithm description is not present at all.}
Fix the initial partition
$P_I :=\{ I_1, \dots, I_{\ell}\}$ and \OPT's final partition
$P_F := \{F_1, \dots, F_{\ell}\}$.
The \emph{distance} of a~partition
 $P = \{C_1, \dots, C_{\ell}\}$ from the initial partition,
defined as 
$\Delta(P) := \sum_{j=1}^{\ell} | C_j \setminus I_j |$,
 is the number of nodes in $P$ that do not reside in their initial cluster.
In other words,
at least $\Delta(P)$ node migrations are required in order to reach the partition $P$ from $P_I$, and thus
$\OPT \geq \Delta^*:= \Delta(P_F) $ \maciek{Confusing. We have space. First define :=, then say that OPT bounds}.
%We  use $ \Delta(P_F) $ wherever the initial partition is clear from the context
The scheme of the algorithm can be found in the  appendix (Algorithm \ref{alg:ppl}).

With each repartitioning \maciek{Unclear what repartitioning means. Maybe with each merge action?},
\PPL moves to a~component respecting partition that minimizes the distance to the initial partition $P_I$.
As a~result,
\PPL never moves to a~partition that is more than $\Delta^*$  node migrations away from $P_I$.
This invariant latter ensures us that \PPL does not pay too much while recovering $P_F$.
We emphasize that a~repartitioning by \PPL replaces the current partition $P$ with a~perfect partition closest to $P_I$.
This way \PPL never moves to a~partition beyond the distance $\Delta^*$.

@@

\begin{theorem}	\label{thm:upperbound}
	\PPL reaches the final partition $P_F$
	 and it is $(2\cdot (k-1)\cdot\ell)$-competitive.
\end{theorem}
\begin{proof}
	\maciek{Maybe: "On each inter-cluster request, the algorithm enumerates all component respecting $\ell$-way partitions in the order of increasing distance to $P_I$" Then we don't need to say "i.e." and two sentences that explain the same thing}
	On each inter-cluster request,
	the algorithm enumerates all component respecting $\ell$-way partitions of components
	that are in the same (closest) distance to $P_I$.
	That is, 
	once it reaches a~partition $P$ at distance $\Delta^* = \Delta(P)$,
	it does not move to a~partition
	$P', \Delta(P') > \Delta^*$,
	before it enumerates all partitions at distance $\Delta^*$.
	Therefore,
	\PPL eventually reaches  the partition $P_F$ at distance $\Delta^*=\OPT$.
	With each distinct request, 
	the size of some component increases by one.
	For any cluster $F_i \in P_F$,
	we have $\sum_{C \in F_i} |C| = k$.
	A component $C \in F_i$,
	 initially begins as an isolated node and it grows by gaining $|C|-1$ more nodes.
	Hence, the total number of times a~component in $F_i$ grows is 
	$\sum_{C \in F_i} (|C|-1) \leq k-1$.
	Therefore, there are at most  $(k-1)\cdot\ell $ distinct requests
	for which \PPL performs a~repartitioning and
	 \PPL performs at most $(k-1)\cdot\ell $ repartitions.
	By Lemma \ref{lemma:rebalancecost},
	each repartitioning costs at most $2\cdot\OPT$.
	The total cost is thus at most $2\cdot\OPT\cdot (k-1) \cdot\ell$, which implies the competitive ratio.
\end{proof}

@@

\subsection{Improved Algorithm for Online Rematching} \label{sec:k2}
In this section,
we present \RM,
 an algorithm for clusters of capacity $k=2$.
We interpret a~pair of nodes collocated in one cluster as a~``matched'' pair.
Hence,
the problem is an online variant of the maximal matching problem where
%a match is retractable.
%That is, 
a matched pair can separate in order to ``rematch'' with two other nodes.
Rematching is necessary for  maximizing intra-cluster communications,
which is dual \maciek{dual has specific meaning. Is it correct here? I think you meant "equivalent" here} to minimizing inter-cluster communications.
This is known as the  \emph{online rematching} 
problem and a~non-strict 7-competitive algorithm is already given by~\cite{repartition-disc},
in which the ratio comes with an additive factor $O(\alpha\ell^2)$.
We do not only improve upon their competitive ratio,
but also show that our ratio holds \emph{strictly}
(i.e., with no additive factor).
Our algorithm is slightly simpler than the one in~\cite{repartition-disc}, 
while our analysis is significantly  simpler and more concise,
thanks to the charging scheme we devise here \maciek{Way too strong. At best, analysis is slightly simpler. Just compare volumes and the notation needed.}.

@@

Consider two  pairs that share the same node, 
i.e.~\emph{intersecting pairs},
and the set of requests that causes the collocation of both pairs,
at some  times during  $[1,m]$.
%i.e.~the entire duration of the sequence.
Observe that \OPT must pay a~non-zero cost
for these requests over the entire $\sigma$,
since it cannot have both pairs initially collocated.
However,
we can charge the whole cost \maciek{Whole cost unclear, I think} to \OPT only the first time \RM collocates a~pair,
and not at any consequent time when \RM collocates it a~second time.
Otherwise,
 \OPT is possibly charged for the same cost repeatedly.
For this reason,
we charge \OPT a~cost inflicted by a~pair,
if and only if  \OPT incurs that cost after the latest separation of the pair by \RM.
%%

\begin{proof}[Proof of Theorem \ref{thm:k=2}]
	Fix an input sequence of requests
$\sigma:=\{\sigma_1,\dots, \sigma_m\}$.
	Assume that \RM collocates a pair $\set{u,v}$ at time $t$.
	The value of $C_{\set{u,v}}$ at $t$,
	denoted $C^t_{\set{u,v}}$, 
	reaches $\lambda$ immediately before \RM resets the counter.
	For any interval $[t_1,t_2]$, by $\sigma_{\set{x,y}}[t_1,t_2]$ we denote the set of all requests to a~pair $\set{x,y}$ that arrive
	during $[t_1,t_2]$.
	%and its overall cost to \OPT by
	%$\OPT (\sigma_{\{u,v\}}[t_1,t_2])$.
	We may use $\sigma_{\set{x,y}}$ whenever
	 the interval $[t_1,t_2]$ is clear from the context \maciek{As a reader I would be horrified seeing this sentence, expecting shenanigans. Better to fix it in the paragraph}.
	%
	%Else, this is not the first time and
	%Alg collocates $\{u,v\}$ also some time prior to $t$.

@@

	Next,
	we bound  costs incurred by \OPT for requests that trigger collocation of $\{u,v\}$ at $t \in T$.
	If $t$ is the first time that \RM collocates $\{u,v\}$,
	 then  \OPT pays
	$\lambda$ for serving requests in $\sigma_{\set{u,v}}[0,t]$ (remotely),
	or $\alpha$ for collocating the pair and
	serving (some of) the requests with  cost zero.
	Therefore in this case,
	$\OPT(\sigma_{\{u,v\}}(0,t]) \geq  \min{\{ \lambda,\alpha \}}$.
	%
	Otherwise,
	it is not the first collocation and
	consider times $t'$ and $t''$ as defined previously,
	 and let 
	$R_t := \sigma_{\{u,w\}}(t'',t'] \cup \sigma_{\{u,v\}}(t',t] $.
	We define $R_{t'}$ for the collocation at $t'$  analogously.
	Then,
	$\OPT(R_t) =\OPT (\sigma_{\{u,w\}}) 
	+ \OPT(\sigma_{\{u,v\}}) $.
	If \OPT has both pairs separated during their respective intervals,
	then obviously it pays $2\lambda$ during those intervals.
	Note that \OPT cannot have both pairs collocated at the same time.
%	However, it can have one of the pairs already collocated prior to its respective interval.
	Let us assume \OPT has one of the pairs,
	e.g.~$\{u,v\}$,
	 collocated already prior its respective interval, $(t',t]$,
	 and keeps it so during the interval.
	 Then it pays zero while serving $\sigma_{\{u,v\}}$.
	Hence,
	it must pay $\alpha$ for collocating the other pair, in this case $\{u,w\}$,
	or (resp., and) it pays (resp., up to) $\lambda$ for serving (resp., some of) requests in $\sigma_{\{u,w\}}$ \maciek{Rewrite without these resp.s}. 
	Therefore in any case,
	$\OPT(R_t) \geq  \min{\{ \lambda,\alpha \}} = \alpha$.

	It remains to bound the cost  incurred by \OPT due to requests to $\{u,v\}$ that do not lead to its collocation until the end of the sequence at $t=m$.
	We bound the cost analogously to the case where \RM collocates $\{u,v\}$.
	If $\{u,v\}$ is not collocated in the initial matching
	and \RM never collocates it,
	then $ C^{m}_{ \{u,v\} } =| \sigma_{\{u,v\}}[1,m] |$.
	\OPT pays
	$\OPT(\sigma_{\{u,v\}}[1,m]) 
	\geq \min{ \{ \alpha, C^{m}_{ \{u,v\} } \} }$,
	for collocating this pair or (and) paying for (resp.~some of) requests in $\sigma_{\{u,v\}}[1,m]$.
	Else,
	either $\{u,v\}$ is collocated in the initial matching
	or \RM collocates it at some point.
	Then there exists an intersecting pair $\set{u,w}$
	that is collocated by \RM at $t' < m$,
	separating $\{u,v\}$.
	We define times $t'' < t' < m$ similarly to the former case \maciek{doing thing "similarly" is a trap zone in proofs, and we should avoid even suggesting it. I think that here you mean defining exactly the same.}.
	Let $R^*_{\{u,v\}} := \sigma_{\{u,w\}} (t'',t'] \cup \sigma_{\{u,v\}} (t',m]$.
	Then,
	\OPT must pay for collocating at least one pair or (and) serving requests 
	to the other pair remotely.
	Thus,
	$\OPT (R^*_{\{u,v\}}) 
	\geq  \min{ \{ C^{m}_{ \{u,v\}}, \alpha \}}$.

>>>>

	We proceed in similar steps by joining the current pivot node to a ground set of the same origin residing in a different cluster.
	Consider the step $i$,
	when the isolated node $x_i$ is collocated with $B$.
	 We issue a request between $x_i$ and some node in $C_i$,
	 where $C_i$ is the largest ground set
	 s.t.~$I(C_i) = I(x_i), C_i \neq \{x_0,y_0\}$.
	 Then \ALG must collocate the new ground set
	  $\{x_i\} \cup C_i $ in one cluster.
%	 only if  a feasible partition exists
%	 that collocates all nodes in $\{x_i\} \cup C_i $.
	Any feasible partition replaces $x_i$ with some isolated node $x_{i+1}$,
	as the new ground set $\{x_i\} \cup C_i $  may not be ever split.
%	Otherwise,
%	no such feasible partition exists and
%	we terminate the process.
	We terminate the process once the number of remaining isolated nodes 
	is less than $\ell+3$.
	At each step $i$,
	the number of isolated nodes decrease either by one,
	 or by two if $C_i$ is a singleton.
	Therefore, once the process terminates,
	in any case at least $\ell+1$ isolated nodes are left.

@@

We maintain subsets of frequently communicating nodes as \emph{components}.
Initially,
each node constitutes a single-node component which we refer to as a \emph{singleton} component, 
and the node in such component is an \emph{isolated} node.
We define larger components in terms of smaller components.
Concretely,
given a (sub)sequence of requests,
two components $C_1$ and $C_2$
 \emph{merge} into one component as soon as
 for some pair of nodes $v_1 \in C_1$ and $v_2 \in C_2$,
the frequency of requests $\{v_1,v_2\}$ reaches a certain threshold through the sequence.
We keep all nodes of a~component always collocated in the same cluster,
i.e., when we  move a~node,
we move the whole component that contains it.
A~partition that has every component collocated 
is a~\emph{component respecting} partition.
%Intuitively, a component preserves the history of requests between its nodes.

@@

The algorithm in this section and the algorithm for $k=3$ (cf. Section~\ref{sec:k3}) are modified versions of the algorithm DET from~\cite{repartition-disc}.
The difference is in the choice of partition after a~component merge.
In DET, the partition was arbitrary.
In the algorithm for the learning model,
we choose a~component respecting partition closest to the initial partition.
In the algorithm from Section~\ref{sec:k3}, we choose the partition closest to the current partition (a repartition of minimum cost).
%A generic sketch of these two algorithms can be found in the appendix (Algorithm \ref{alg:ppl}).

\noindent
\textbf{Perfect Partition Learner algorithm.}
Fix the initial partition
$P_I :=\{ I_1, \dots, I_{\ell}\}$ and \OPT's final partition
$P_F := \{F_1, \dots, F_{\ell}\}$.
The \emph{distance} of a~partition
 $P = \{C_1, \dots, C_{\ell}\}$ from the initial partition,
defined as 
$\Delta(P) := \sum_{j=1}^{\ell} | C_j \setminus I_j |$,
 is the number of nodes in $P$ that do not reside in their initial cluster.
In other words,
at least $\Delta(P)$ node migrations are required in order to reach the partition $P$ from $P_I$, and thus
$\OPT \geq \Delta(P_F) $.

Now we describe the algorithm \PPL.
On each inter-cluster request $\{u,v\}$, 
\PPL creates a new components by merging
 the two components that contain nodes $u$ and $v$.
In order to collocate nodes of the new component,
\PPL moves to a~component respecting partition that minimizes the distance to the initial partition $P_I$.
The scheme of the algorithm can be found in the  appendix (Algorithm \ref{alg:ppl}).

\PPL never moves to a~partition that is more than $ \Delta^*:= \Delta(P_F) $   migrations away from $P_I$.
This invariant latter ensures us that \PPL does not pay too much while recovering $P_F$.
We emphasize that a~\emph{repartitioning} by \PPL replaces the current partition $P$ with a~perfect partition closest to $P_I$.
This way \PPL never moves to a~partition beyond the distance $\Delta^*$.

@@

\begin{theorem}	\label{thm:upperbound}
	\PPL reaches the final partition $P_F$
	 and it is $(2\cdot (k-1)\cdot\ell)$-competitive.
\end{theorem}
\begin{proof}
	On each inter-cluster request,
	the algorithm enumerates all component respecting $\ell$-way partitions of components
	that are in the same (closest) distance to $P_I$.
	That is, 
	once it reaches a~partition $P$ at distance $\Delta^* = \Delta(P)$,
	it does not move to a~partition
	$P', \Delta(P') > \Delta^*$,
	before it enumerates all partitions at distance $\Delta^*$.
	Therefore,
	\PPL eventually reaches  the partition $P_F$ at distance $\Delta^*=\OPT$.
	With each distinct request, 
	the size of some component increases by one.
	For any cluster $F_i \in P_F$,
	we have $\sum_{C \in F_i} |C| = k$.
	A component $C \in F_i$,
	 initially begins as an isolated node and it grows by gaining $|C|-1$ more nodes.
	Hence, the total number of times a~component in $F_i$ grows is 
	$\sum_{C \in F_i} (|C|-1) \leq k-1$.
	Therefore, there are at most  $(k-1)\cdot\ell $ distinct requests
	for which \PPL performs a~repartitioning and
	 \PPL performs at most $(k-1)\cdot\ell $ repartitions.
	By Lemma \ref{lemma:rebalancecost},
	each repartitioning costs at most $2\cdot\OPT$.
	The total cost is thus at most $2\cdot\OPT\cdot (k-1) \cdot\ell$, which implies the competitive ratio.
\end{proof}

@@

\subsection{Improved Algorithm for Online Rematching} \label{sec:k2}
In this section,
we present \RM,
 an algorithm for clusters of capacity $k=2$.
We interpret a~pair of nodes collocated in one cluster as a~``matched'' pair.
Hence,
the problem is an online variant of the maximal matching problem where
%a match is retractable.
%That is, 
a matched pair can separate in order to ``rematch'' with two other nodes.
Rematching is necessary for  maximizing intra-cluster communications,
which is equivalent to minimizing inter-cluster communications.
This is known as the  \emph{online rematching} 
problem and a~non-strict 7-competitive algorithm is already given by~\cite{repartition-disc},
in which the ratio comes with an additive factor $O(\alpha\ell^2)$.
We do not only improve upon their competitive ratio,
but also show that our ratio holds \emph{strictly}
(i.e., with no additive factor).
Our algorithm is slightly simpler than the one in~\cite{repartition-disc}, 
while our analysis is significantly simpler and more concise,
thanks to the charging scheme we devise here.

@@

Consider two  pairs that share the same node, 
i.e.~\emph{intersecting pairs},
and the set of requests that causes the collocation of both pairs,
at some  times during  $[1,m]$.
%i.e.~the entire duration of the sequence.
Observe that \OPT must pay a~non-zero cost
for these requests over the entire $\sigma$,
since it cannot have both pairs initially collocated.
However,
we can charge this cost to \OPT only the first time \RM collocates a~pair,
and not at any consequent time when \RM collocates it a~second time.
Otherwise,
 \OPT is possibly charged for the same cost repeatedly.
For this reason,
we charge \OPT a~cost inflicted by a~pair,
if and only if  \OPT incurs that cost after the latest separation of the pair by \RM.
%%

\begin{proof}[Proof of Theorem \ref{thm:k=2}]
	Fix an input sequence of requests
$\sigma:=\{\sigma_1,\dots, \sigma_m\}$.
	Assume that \RM collocates a pair $\set{u,v}$ at time $t$.
	The value of $C_{\set{u,v}}$ at $t$,
	denoted $C^t_{\set{u,v}}$, 
	reaches $\lambda$ immediately before \RM resets the counter.
	For any interval $[t_1,t_2]$, by $\sigma_{\set{x,y}}[t_1,t_2]$ we denote the set of all requests to a~pair $\set{x,y}$ that arrive
	during $[t_1,t_2]$.
	%and its overall cost to \OPT by
	%$\OPT (\sigma_{\{u,v\}}[t_1,t_2])$.
	We may use $\sigma_{\set{x,y}}$ whenever
	 the interval $[t_1,t_2]$ is clear from the context.
	%
	%Else, this is not the first time and
	%Alg collocates $\{u,v\}$ also some time prior to $t$.

@@

	Next,
	we bound  costs incurred by \OPT for requests that trigger collocation of $\{u,v\}$ at $t \in T$.
	If $t$ is the first time that \RM collocates $\{u,v\}$,
	 then  \OPT pays
	$\lambda$ for serving requests in $\sigma_{\set{u,v}}[0,t]$ (remotely),
	or $\alpha$ for collocating the pair and
	serving (some of) the requests with  cost zero.
	Therefore in this case,
	$\OPT(\sigma_{\{u,v\}}(0,t]) \geq  \min{\{ \lambda,\alpha \}}$.
	%
	Otherwise,
	it is not the first collocation and
	consider times $t'$ and $t''$ as defined previously,
	 and let 
	$R_t := \sigma_{\{u,w\}}(t'',t'] \cup \sigma_{\{u,v\}}(t',t] $.
	We define $R_{t'}$ for the collocation at $t'$  analogously.
	Then,
	$\OPT(R_t) =\OPT (\sigma_{\{u,w\}}) 
	+ \OPT(\sigma_{\{u,v\}}) $.
	If \OPT has both pairs separated during their respective intervals,
	then obviously it pays $2\lambda$ during those intervals.
	Note that \OPT cannot have both pairs collocated at the same time.
%	However, it can have one of the pairs already collocated prior to its respective interval.
	Let us assume \OPT has one of the pairs,
	e.g.~$\{u,v\}$,
	 collocated already prior its respective interval, $(t',t]$,
	 and keeps it so during the interval.
	 Then it pays zero while serving $\sigma_{\{u,v\}}$.
	Hence,
	it must pay $\alpha$ for collocating the other pair, in this case $\{u,w\}$,
	or (resp., and) it pays (resp., up to) $\lambda$ for serving (resp., some of) requests in $\sigma_{\{u,w\}}$. 
	Therefore in any case,
	$\OPT(R_t) \geq  \min{\{ \lambda,\alpha \}} = \alpha$.

	It remains to bound the cost  incurred by \OPT due to requests to $\{u,v\}$ that do not lead to its collocation until the end of the sequence at $t=m$.
	We bound the cost analogously to the case where \RM collocates $\{u,v\}$.
	If $\{u,v\}$ is not collocated in the initial matching
	and \RM never collocates it,
	then $ C^{m}_{ \{u,v\} } =| \sigma_{\{u,v\}}[1,m] |$.
	\OPT pays
	$\OPT(\sigma_{\{u,v\}}[1,m]) 
	\geq \min{ \{ \alpha, C^{m}_{ \{u,v\} } \} }$,
	for collocating this pair or (and) paying for (resp.~some of) requests in $\sigma_{\{u,v\}}[1,m]$.
	Else,
	either $\{u,v\}$ is collocated in the initial matching
	or \RM collocates it at some point.
	Then there exists an intersecting pair $\set{u,w}$
	that is collocated by \RM at $t' < m$,
	separating $\{u,v\}$.
	We define times $t'' < t' < m$ analogously to the former case.
	Let $R^*_{\{u,v\}} := \sigma_{\{u,w\}} (t'',t'] \cup \sigma_{\{u,v\}} (t',m]$.
	Then,
	\OPT must pay for collocating at least one pair or (and) serving requests 
	to the other pair remotely.
	Thus,
	$\OPT (R^*_{\{u,v\}}) 
	\geq  \min{ \{ C^{m}_{ \{u,v\}}, \alpha \}}$.
